---
title: "Parametric_Regression"
author: "yc4384_Yangyang_Chen"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(survminer)
library(tidyverse)
```


```{r, include = FALSE}
data("pbc", package = "survival")
# Inspect the dataset
str(pbc)
skimr::skim(pbc)

pbcseq_cleaned <- pbc %>%
  filter(trt != "NA", 
         chol != "NA",
         trig != "NA",
         platelet != "NA",
         copper != "NA") %>%
  mutate(status = case_when(
    status == 0 ~ 0, 
    status == 1 ~ 1, 
    status == 2 ~ 1   
  ),
 trt = case_when(
 trt == 1 ~ 1,
 trt == 2 ~ 0)) %>%
  mutate(
    sex = as.factor(sex),
    trt = as.factor(trt),
    ascites = as.factor(ascites),
    edema = as.factor(edema),
    hepato = as.factor(hepato),
    spiders = as.factor(spiders),
    stage = as.factor(stage)
  )

skimr::skim(pbcseq_cleaned)
pbcseq_cleaned$surv_object <- Surv(time = pbcseq_cleaned$time, event = pbcseq_cleaned$status)
```

## Failed of Log-Rank Tests
```{r, echo = FALSE}
# Log-rank test (rho = 0)
logrank_test <- survdiff(surv_object ~ trt, data = pbcseq_cleaned , rho = 0)
logrank_test
```

Why we should consider some alternative approaches based on parametric models:

The assumption of proportional hazards might not be appropriate (based on major departures).

## I. Exponential Regression

Assume \(T_i\) follows an exponential distribution with a parameter \(\lambda\) that depends on \(\mathbf{Z}_i\), say \(\lambda_i = \Psi(\mathbf{Z}_i)\). Then we can write:

\[
T_i \sim \text{exponential}(\Psi(\mathbf{Z}_i))
\]

### (1) Fit Exponential Regression model by Stepwise Selection

#### a. Multivariate Analysis

- Fit a full model with all candidate variables, then use bidirectional stepwise selection to identify the optimal subset of predictors based on AIC.

```{r, echo = FALSE}
# Stepwise selection with all variables
full_model <- survreg(Surv(time, status) ~ trt + spiders+ edema + bili + albumin + copper + protime + stage + age + sex + ascites + hepato,
                      data = pbcseq_cleaned,
                      dist = "exponential")
selected_model <- step(full_model, direction = "both", trace = FALSE)

# Summarize selected model
summary(selected_model)
```

#### b. Fit Exponential model

```{r, echo = FALSE}
exp_model <- survreg(Surv(time, status) ~ bili + albumin + copper + protime + stage + sex, 
                         data = pbcseq_cleaned, 
                         dist = "exponential")
summary(exp_model)
```

### (2) Perform Likelihood Ratio Test

```{r, echo = FALSE}
# Perform Likelihood Ratio Test
null_model <- survreg(Surv(time, status) ~ 1, 
                      data = pbcseq_cleaned, 
                      dist = "exponential")

anova(null_model, exp_model, test = "Chisq")
```

Interpretation of Results:

	-	Model Significance: The large deviance difference $(\Delta D = 124.192)$ and small p-value $( p < 0.05 )$ indicate that the exponential model with the predictors (bili, albumin, copper, protime, stage, and sex) provides a significantly better fit to the data than the null model.

  -	Predictors Significance: The predictors in the exponential model contribute significantly to explaining the survival times, as evidenced by the highly significant p-value.
	
	- Model Fit: The exponential model improves the fit over the null model by reducing deviance, which measures unexplained variation in the data.
	
### (3) Model Fit Comparison using AIC

```{r, echo = FALSE}
# Model Fit Comparison using AIC
AIC(null_model, exp_model)

# Diagnostic Plot: Comparing Model Survival to Kaplan-Meier
km_fit <- survfit(Surv(time, status) ~ 1, data = pbcseq_cleaned)
plot(km_fit, xlab = "Time", ylab = "Survival Probability", main = "Kaplan-Meier vs Exponential")
lines(seq(0, max(pbcseq_cleaned$time), by = 10),
      exp(-exp_model$scale * seq(0, max(pbcseq_cleaned$time), by = 10)), 
      col = "red", lwd = 2)
legend("topright", legend = c("Kaplan-Meier", "Exponential Model"), col = c("black", "red"), lty = 1)
```

Interpretation of Results:

- The issue in the Kaplan-Meier vs Exponential plot seems to be that the Exponential Model line (red) is a constant hazard (flat line) and does not fit the Kaplan-Meier curve (black) well. This happens because:

  1.	Exponential Model Assumption: The exponential distribution assumes a constant hazard rate over time, which may not match the actual survival pattern in your dataset.
  2.	Poor Fit: The Kaplan-Meier curve indicates non-constant hazard rates (e.g., survival probabilities decrease differently over time), which suggests that the exponential model may not be appropriate for this data.
	
- Consider Alternative Parametric Models:

  - The Weibull model is a more flexible parametric survival model that allows for non-constant hazard rates. If the hazard rate varies over time, the Weibull model may provide a better fit.

## II. Welbull Regression

Weibull Survival Function:

- The Weibull survival function is calculated as:

$$S(t) = \exp\left(-\left(\frac{t}{\lambda}\right)^k\right)$$

### (1) Fit Weibull model by Stepwise Selection

#### a. Multivariate Analysis:

- Fit a full model with all candidate variables, then use bidirectional stepwise selection to identify the optimal subset of predictors based on AIC.

```{r, echo = FALSE}
# Stepwise selection with all variables
full_model <- survreg(Surv(time, status) ~ trt + spiders+ edema + bili + albumin + copper + protime + stage + age + sex + ascites + hepato, data = pbcseq_cleaned, dist = "weibull")
selected_model <- step(full_model, direction = "both", trace = FALSE)

# Summarize selected model
summary(selected_model)
```

#### b. Fit Weibull Regression Model

```{r, echo = FALSE}
weibull_model <- survreg(Surv(time, status) ~ edema + albumin + protime + stage + sex, data = pbcseq_cleaned, dist = "weibull")	
summary(weibull_model)
```

### (2) Perform Likelihood Ratio Test

```{r, echo = FALSE}
# Perform Likelihood Ratio Test
null_model <- survreg(Surv(time, status) ~ 1, 
                      data = pbcseq_cleaned, 
                      dist = "weibull")

anova(null_model, exp_model, test = "Chisq")
```

Interpretation of Results:

	-	Model Significance: The large deviance difference $(\Delta D = 119.1457)$ and small p-value $( p < 0.05 )$ indicate that the weibull model with the predictors (bili, albumin, copper, protime, stage, and sex) provides a significantly better fit to the data than the null model.

  -	Predictors Significance: The predictors in the weibull model contribute significantly to explaining the survival times, as evidenced by the highly significant p-value.
	
	- Model Fit: The exponential model improves the fit over the null model by reducing deviance, which measures unexplained variation in the data.
	
### (3) Diagnostic Plot

```{r, echo = FALSE}
# Extract Weibull parameters
scale_weibull <- exp(coef(weibull_model)["(Intercept)"])  # Scale parameter (λ)
shape_weibull <- 1 / weibull_model$scale  # Shape parameter (k)

# Generate time points for Weibull survival function
time_points <- seq(0, max(pbcseq_cleaned$time, na.rm = TRUE), by = 10)

# Calculate Weibull survival probabilities
weibull_survival <- exp(-((time_points / scale_weibull) ^ shape_weibull))

# Plot Kaplan-Meier
km_fit <- survfit(Surv(time, status) ~ 1, data = pbcseq_cleaned)
plot(km_fit, xlab = "Time", ylab = "Survival Probability", main = "Kaplan-Meier vs Weibull Model")

# Add Weibull survival curve
lines(time_points, weibull_survival, col = "blue", lwd = 2)

# Add legend
legend("bottomleft", legend = c("Kaplan-Meier", "Weibull Model"), col = c("black", "blue"), lty = 1)
```

Interpretation of the Plot

	1.	Model Fit: The Weibull model generally provides a good fit to the data. Its alignment with the Kaplan-Meier curve in the early and mid survival times indicates that the Weibull assumption is appropriate for much of the dataset.

	2.	Weibull Assumptions: The plot supports the assumption of a Weibull distribution for most of the data. Further diagnostics (e.g., log-log plots, residual analysis) can confirm this.
	
## III. Parametric Models Comparison

### (1) AIC Comparisons

- To assess whether the Weibull model provides a significantly better fit than the exponential model, compare their AIC values:

```{r, echo = FALSE}
AIC(exp_model, weibull_model)
```

Interpretation:

While the exponential model achieves a slightly lower AIC, the Weibull model is slightly more flexible and is therefore the preferred choice based on these results.

### (2) Graphical Diagnostics

- Plot the log cumulative hazard $(log(-log(S(t))))$ against the log of time. If the points form a straight line, it confirms that the Weibull model is appropriate:

```{r, echo = FALSE}
km_fit <- survfit(Surv(time, status) ~ 1, data = pbcseq_cleaned)
plot(log(km_fit$time), log(-log(km_fit$surv)), xlab = "log(Time)", ylab = "log(-log(Survival))", main = "Log-Log Plot")
abline(a = coef(weibull_model)["(Intercept)"], b = 1 / weibull_model$scale, col = "blue")
```

Interpretation:

In the log-log plot, the points approximately form a straight line, it suggests that the Weibull distribution is a good fit for the data. Minor deviations in the tails may warrant further exploration, but the Weibull model appears to capture the main survival patterns effectively. 

## IV. Model Diagnostics

### (1) Residual diagnostics

- Residual diagnostics help assess whether the Weibull model fits the survival data well. For survival models, we commonly use Deviance residuals and Cox-Snell residuals.

#### a. Deviance Residuals

- Deviance residuals can be computed for parametric models fitted with `survreg()`:

```{r, echo = FALSE}
# Compute deviance residuals
deviance_resid <- residuals(weibull_model, type = "deviance")

# Plot deviance residuals
plot(pbcseq_cleaned$time, deviance_resid, xlab = "Time", ylab = "Deviance Residuals",
     main = "Deviance Residuals vs Time", pch = 20)
abline(h = 0, col = "red", lty = 2)
```

Interpretation:

-	Deviance residuals scatter randomly around 0, indicating good model fit.

#### b. Cox-Snell Residuals

- Cox-Snell residuals are used to assess overall goodness-of-fit for parametric models. They should follow an exponential distribution with a mean of 1 if the model fits well.

```{r, echo = FALSE}
# Compute Cox-Snell residuals
cox_snell_resid <- pbcseq_cleaned$time * exp(-predict(weibull_model, type = "lp"))

# Fit a Kaplan-Meier survival curve to the Cox-Snell residuals
km_cox_snell <- survfit(Surv(cox_snell_resid, pbcseq_cleaned$status) ~ 1)

# Plot cumulative hazard of Cox-Snell residuals
plot(km_cox_snell$time, -log(km_cox_snell$surv), type = "s", xlab = "Cox-Snell Residuals",
     ylab = "Cumulative Hazard", main = "Cox-Snell Residuals")
abline(0, 1, col = "red", lty = 2)  # Add 45-degree line
```

Interpretation:

- The cumulative hazard of the Cox-Snell residuals lie close to the 45-degree line (red line), indicating the Weibull model fits well.

### (2) Goodness-of-fit tests

#### a. Likelihood Ratio Test

- Use the anova() function to compare the Weibull model to a simpler model -- the exponential model:

```{r, echo = FALSE, }
# Perform likelihood ratio test
anova(exp_model, weibull_model, test = "Chisq")
```

Interpretation:

- While the exponential model has a slightly lower deviance (2288.371) compared to the Weibull model (2291.133), the Weibull model is more flexible and is therefore the preferred choice based on these results.

#### b. Compare AIC values

- To assess whether the Weibull model provides a significantly better fit than the exponential model, compare their AIC values:

```{r, echo = FALSE}
AIC(exp_model, weibull_model)
```
Interpretation:

- While the exponential model achieves a slightly lower AIC, the Weibull model is more flexible and is therefore the preferred choice based on these results.